Type-safe client-server communication for functional web applications

### 2014-02-28 paper.tex
Fix Peter's feedback.

To put make the problem a bit more concrete, let's take a look at an example.
Let's say what we want to implement a simple ``chatbox'' component for a
website, to allow visitors to discuss the site's content in real time. Using
mainstream development practives and recent technologies such as WebSockets,
we may come up with something like the program in figure
\ref{lst:javascript-client}.

@@

\begin{itemize}
  \item The programming model is synchronous, giving the programmer a simple,
        linear view of the program flow, eliminating the need to program with
        callbacks and continuations.
  \item Side-effecting code is explicitly designated to run on either the
        client or the server using the type system while pure code can be
        shared by both. Additionally, both client and server code may lift
        general IO computations, allowing for safe IO code reuse within the
        confines of the client or server designated code sections.
  \item Client-server network communication is handled through statically typed
        RPC function calls, extending the reach of Haskell's type checker over
        the network and giving the programmer advance warning when she uses
        network services incorrectly or forgets to update communication code
        as the application's internal protocol changes.
  \item Our model takes the view that the client side is the main driver when
        developing web applications and accordingly delegates the server to
        the role of a computational and/or storage resource, tasked with
        servicing client requests rather than driving the program. While it is
        entirely possible to implement a server-to-client communication channel
        on top of our model, we believe that choosing one side of the
        heterogenous client-server relation as the master helps keeping the
        program flow linear and predictable.
  \item The implementation is built as a library on top of the GHC and Haste
        Haskell compilers, requiring little to no compiler support. Programs
        are compiled twice; once with Haste and once with GHC, to produce the
        final client and server side code respectively.
\end{itemize}

@@

This leaves us choosing between Clean and Haste for our client compiler. Since
most of our abstractions, both the cooperative multitasking and the type-safe
client-server communication itself, are easily expressed as monads, well
supported in Haskell thanks to do-notation, the choice fell to Haste.

@@

Functions are represented as plain Javascript functions, one of the blessings
of targeting a high level language, and application is can therefore be
implemented as its Javascript counterpart in most cases. In the general case,
however, functions may be curried. For such cases where the arity of an applied
function can not be determined statically, application is implemented using the
eval/apply method described in\ \cite{fastcurry} instead.

>>>>

To put make the problem a bit more concrete, let's take a look at an example.
Let's say that we want to implement a simple ``chatbox'' component for a
website, to allow visitors to discuss the site's content in real time. Using
mainstream development practives and recent technologies such as WebSockets,
we may come up with something like the program in figure
\ref{lst:javascript-client}.

@@

\begin{itemize}
  \item The programming model is synchronous, giving the programmer a simple,
        linear view of the program flow, eliminating the need to program with
        callbacks and continuations.
  \item Side-effecting code is explicitly designated to run on either the
        client or the server using the type system while pure code can be
        shared by both. Additionally, general IO computations may be lifted
        into both client and server code, allowing for safe IO code reuse
        within the confines of the client or server designated code sections.
  \item Client-server network communication is handled through statically typed
        RPC function calls, extending the reach of Haskell's type checker over
        the network and giving the programmer advance warning when she uses
        network services incorrectly or forgets to update communication code
        as the application's internal protocol changes.
  \item Our model takes the view that the client side is the main driver when
        developing web applications and accordingly delegates the server to
        the role of a computational and/or storage resource, tasked with
        servicing client requests rather than driving the program. While it is
        entirely possible to implement a server-to-client communication channel
        on top of our model, we believe that choosing one side of the
        heterogenous client-server relation as the master helps keeping the
        program flow linear and predictable.
  \item The implementation is built as a library on top of the GHC and Haste
        Haskell compilers, requiring little to no compiler support. Programs
        are compiled twice; once with Haste and once with GHC, to produce the
        final client and server side code respectively.
\end{itemize}

@@

This leaves us choosing between Clean and Haste for our client compiler. Most
of our abstractions, both the cooperative multitasking and the type-safe
client-server communication itself, are easily expressed as monads, well
supported in Haskell thanks to do-notation. Add the fact that one of this
work's authors is the original author of Haste, giving us a better position to
troubleshoot any code based thereupon, and Haste becomes the better choice.

@@

Functions are represented as plain Javascript functions, one of the blessings
of targeting a high level language, and application can therefore be
implemented as its Javascript counterpart in most cases. In the general case,
however, functions may be curried. For such cases where the arity of an applied
function can not be determined statically, application is implemented using the
eval/apply method described in\ \cite{fastcurry} instead.
